---
title: "UX, Accessibility, and Mobile Library Services"
---

```{r}
#| label: setup
#| include: false

latex_escape <- function(x) {
  x <- gsub("\\", "<<<BACKSLASH>>>", as.character(x), fixed = TRUE)
  x <- gsub("([#$%&_{}])", "\\\\\\1", x, perl = TRUE)
  x <- gsub("~", "\\textasciitilde{}", x, fixed = TRUE)
  gsub("<<<BACKSLASH>>>", "\\textbackslash{}", x, fixed = TRUE)
}

ict_table <- function(data, caption, label = NULL, widths = NULL) {
  data <- as.data.frame(data, stringsAsFactors = FALSE)
  names(data) <- gsub("_", " ", names(data), fixed = TRUE)
  anchor <- if (!is.null(label)) paste0("{#tbl-", label, "}") else ""
  if (knitr::is_latex_output()) {
    align <- paste(rep("Y", ncol(data)), collapse = "")
    header <- paste(vapply(names(data), latex_escape, character(1)), collapse = " & ")
    rows <- apply(data, 1, function(row) {
      paste(vapply(row, latex_escape, character(1)), collapse = " & ")
    })
    if (grepl("^Key concepts", caption)) {
      column_weights <- if (is.null(widths) || length(widths) != ncol(data)) {
        rep(1, ncol(data))
      } else {
        widths
      }
      column_widths <- 0.94 * column_weights / sum(column_weights)
      long_align <- paste(
        sprintf(">{\\raggedright\\arraybackslash}p{%.3f\\textwidth}", column_widths),
        collapse = ""
      )
      out <- c(
        sprintf("\\begin{longtable}{%s}", long_align),
        sprintf("\\caption{%s}\\label{tbl-%s}\\\\", latex_escape(caption), label),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endfirsthead",
        sprintf("\\multicolumn{%d}{c}{\\tablename\\ \\thetable{} -- continued} \\\\" , ncol(data)),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endhead",
        sprintf("\\midrule \\multicolumn{%d}{r}{Continued on next page} \\\\" , ncol(data)),
        "\\endfoot", "\\bottomrule", "\\endlastfoot",
        paste0(rows, " \\\\"), "\\end{longtable}"
      )
    } else {
      out <- c(
        sprintf("\\begin{tabularx}{\\textwidth}{%s}", align),
        "\\toprule", paste0(header, " \\\\"), "\\midrule",
        paste0(rows, " \\\\"), "\\bottomrule", "\\end{tabularx}"
      )
    }
  } else if (knitr::is_html_output()) {
    if (grepl("^Key concepts", caption)) {
      out <- knitr::kable(data, format = "html", caption = caption, escape = TRUE,
                          table.attr = if (is.null(label)) "" else sprintf("id='tbl-%s'", label))
    } else {
      out <- knitr::kable(data, format = "html", escape = TRUE)
    }
  } else {
    if (grepl("^Key concepts", caption)) {
      out <- c(paste0(": ", caption, " ", anchor), "", knitr::kable(data, format = "pipe"))
    } else {
      out <- knitr::kable(data, format = "pipe")
    }
  }
  knitr::asis_output(paste(out, collapse = "\n"))
}
```

## Chapter overview

A student opens the university library discovery service on a low-cost phone while riding home. The search box fits the screen, but a cookie notice covers the search button. The results page loads several megabytes of images. Filters are hidden behind an unlabeled icon. The student finds an article, yet the sign-in form loses entered text when the connection drops. Another student using a keyboard cannot reach the date filter, while a screen-reader user hears five links all announced as “Read more.” The site is technically online, but the service is not equally usable.

The central problem of this chapter is **how to design and evaluate information services so that people with varied goals, abilities, devices, languages, connections, and circumstances can complete meaningful tasks with dignity and control**. User experience (UX) is not decoration. It includes the complete encounter with a service: discovering that it exists, understanding its scope, navigating it, interpreting metadata, authenticating, obtaining content, recovering from errors, asking for help, and deciding whether to trust the institution.

Accessibility is also not a checklist added after design. It is the requirement that people with disabilities can perceive, operate, understand, and use a service, including through assistive technologies. Mobile design is more than shrinking desktop pages. It requires prioritizing tasks, supporting touch and keyboard interaction, minimizing data and battery costs, tolerating interruption, and making status and recovery visible.

LIS professionals shape UX through metadata labels, search defaults, navigation, content, forms, policies, procurement, support, and assessment. A vendor may write the interface code, but the library remains responsible for the service it offers. This chapter introduces user research, task analysis, information architecture, usability testing, responsive and mobile-first design, plain language, assistive technology, WCAG 2.2, accessibility evaluation, and ethical UX research.

## Learning objectives

After completing this chapter, you should be able to:

1. **Distinguish** user experience, usability, accessibility, inclusive design, and responsive design.
2. **Construct** a realistic task scenario, usability-test protocol, and observation guide for an LIS service.
3. **Evaluate** selected pages and workflows against relevant WCAG 2.2 success criteria using manual and automated checks.
4. **Identify** barriers involving navigation, metadata labels, content, forms, keyboard use, assistive technology, mobile layout, bandwidth, and error recovery.
5. **Prioritize** findings by user impact, frequency, task criticality, reach, and implementation effort.
6. **Recommend** a redesign that addresses evidence without collecting unnecessary personal data or excluding affected users.

## Key concepts

@tbl-ux-key-concepts establishes the working vocabulary. The terms overlap, but they answer different questions.

```{r}
#| label: tbl-ux-key-concepts
#| tbl-cap: "Key concepts in UX, accessibility, and mobile design"
#| results: asis
ux_key_concepts <- data.frame(
  Term = c("User experience", "Usability", "Accessibility", "Inclusive design", "Universal design",
           "User research", "Stakeholder", "Persona", "Task", "Scenario", "Journey map",
           "Information architecture", "Navigation", "Prototype", "Usability test", "Think-aloud",
           "Assistive technology", "Screen reader", "Keyboard access", "Focus indicator", "Semantic HTML",
           "Alternative text", "Caption", "Contrast", "Plain language", "Responsive design", "Mobile first",
           "Breakpoint", "Progressive enhancement", "WCAG", "Success criterion", "Conformance", "Severity"),
  Working_definition = c(
    "A person's complete experience before, during, and after interacting with a product or service.",
    "The extent to which specified users achieve specified goals effectively, efficiently, and satisfactorily in context.",
    "The quality of enabling people with disabilities to perceive, operate, understand, and robustly use a service.",
    "A process that includes varied people and edge conditions in design rather than assuming one typical user.",
    "Design intended to be usable by the widest practical range of people without separate versions.",
    "Systematic study of users, tasks, contexts, barriers, expectations, and outcomes.",
    "A person or group affected by or able to affect a service.",
    "A research-based representation of a user group used to focus design decisions.",
    "A goal-directed activity a person attempts to complete.",
    "A concise context and goal given to a participant without prescribing steps.",
    "A representation of stages, touchpoints, decisions, emotions, evidence, and failures in a service experience.",
    "The organization, naming, relationships, and navigation of information spaces.",
    "Mechanisms that help people know where they are and move to desired content or actions.",
    "A representation built to explore or test a design before full implementation.",
    "Observation of representative participants attempting realistic tasks with a service or prototype.",
    "A method in which participants verbalize what they notice, expect, and decide while attempting a task.",
    "Hardware or software supporting interaction for people with disabilities.",
    "Software that presents interface content through speech, braille, or both.",
    "The ability to reach and use interactive functions without a pointing device.",
    "A visible indication of which element currently receives keyboard input.",
    "HTML elements used according to their meaning so structure and controls are programmatically available.",
    "A text equivalent communicating the purpose or information of a meaningful image.",
    "Synchronized text representing speech and important sound in time-based media.",
    "A measurable difference in luminance or color that supports perceiving text and interface components.",
    "Clear, audience-appropriate communication organized so readers can find and use what they need.",
    "An approach in which layout and components adapt to viewport and device conditions.",
    "A design approach beginning with essential tasks and constrained screens before enhancing larger contexts.",
    "A condition at which responsive presentation rules change.",
    "Providing a reliable core experience first, then adding supported enhancements.",
    "The Web Content Accessibility Guidelines published by W3C.",
    "A testable WCAG requirement at Level A, AA, or AAA.",
    "A claim that identified content meets defined requirements at a stated WCAG level and version.",
    "A reasoned rating of a problem's impact on users and tasks."
  )
)
ict_table(ux_key_concepts, "Key concepts in UX, accessibility, and mobile design",
          "ux-key-concepts", c(1, 3))
```

## Core discussion

### UX is the service, not only the screen

The user interface is the visible and interactive part of a service. UX includes much more. A catalogue search may be easy, but the overall experience fails if holdings are stale, call-number labels are unclear, an elevator is inaccessible, or staff cannot explain an authentication error. A repository may meet technical standards while deposit instructions frighten authors with legal language. A digital exhibit may be visually elegant but omit transcription and provenance.

Usability commonly examines whether intended users can complete tasks effectively, efficiently, and satisfactorily in context. Accessibility examines whether people with disabilities can participate and whether the technology works with varied ways of perceiving and operating it. Inclusive design attends to human diversity and exclusion throughout the design process. These qualities reinforce one another, but one cannot substitute for another. A quick task completion rate does not prove screen-reader access; passing an automated accessibility scan does not prove understandable navigation.

@tbl-ux-questions distinguishes the evidence sought by related practices.

```{r}
#| label: tbl-ux-questions
#| tbl-cap: "Questions and evidence in UX-related evaluation"
#| results: asis
ux_questions <- data.frame(
  Practice = c("UX research", "Usability evaluation", "Accessibility evaluation", "Analytics", "Service assessment"),
  Central_question = c("Who is trying to accomplish what, in which context, and why?",
                       "Can representative users complete defined tasks with acceptable effort and confidence?",
                       "Can people with disabilities perceive, operate, understand, and robustly use the service?",
                       "What aggregate patterns of use and failure appear in instrumented events?",
                       "Does the service produce equitable, mission-aligned outcomes worth sustaining?"),
  Evidence = c("Interviews, observation, diaries, contextual inquiry, support records",
               "Task success, paths, errors, time, comments, observation",
               "Standards review, keyboard test, assistive-technology test, disabled-user testing",
               "Privacy-protected counts, funnels, search patterns, performance measures",
               "Mixed methods, outcomes, costs, reach, equity, policy evidence"),
  Limitation = c("Small samples require careful interpretation", "A task and participant set cannot represent everyone",
                 "Automated tools and standards checks cannot replace lived experience",
                 "Observed events do not explain motivation and may exclude nonusers", "Broad outcomes can be difficult to attribute to one interface")
)
ict_table(ux_questions, "Questions and evidence in UX-related evaluation", "ux-questions", c(1, 2.2, 2.2, 1.7))
```

As @tbl-ux-questions shows, methods answer different questions. Combining them produces a stronger account than declaring that one score represents “the user.” Library usability studies demonstrate the value of observing actual users rather than relying only on professional expectations [@emde2009usability; @vargas2026homepage].

### The evidence-based design cycle

UX work is iterative. Teams frame a service problem, study contexts, model tasks, prototype, evaluate, prioritize, implement, and monitor. Results often reveal that the original problem was incomplete.

```{mermaid}
%%| label: fig-ux-cycle
%%| fig-cap: "An iterative, evidence-based UX and accessibility cycle"
flowchart LR
    F["Frame service problem and equity goal"] --> R["Research users, contexts, tasks, and barriers"]
    R --> M["Model journeys and information architecture"]
    M --> P["Prototype content and interaction"]
    P --> E["Evaluate usability and accessibility"]
    E --> Q["Prioritize by impact, reach, evidence, and effort"]
    Q --> I["Implement and document"]
    I --> O["Monitor outcomes and new exclusion"]
    O --> F
```

As @fig-ux-cycle indicates, accessibility begins during framing and research. If disabled users are invited only after procurement, the team may discover barriers it cannot contractually or technically correct.

### Researching people without inventing a typical user

User research begins with a question, not a favored method. If students cannot find reserve readings, observation and task testing may reveal navigation barriers. If rural users rarely access the repository, interviews may reveal data cost, device sharing, language, or trust issues invisible in analytics.

Useful methods include:

* **interviews** for goals, expectations, language, workarounds, and meaning;
* **contextual inquiry** for observing tasks in their actual setting;
* **diary studies** for experiences distributed over time;
* **surveys** for structured self-report across a larger group;
* **analytics** for aggregate paths and events, with privacy controls;
* **support-log review** for recurrent questions and failures;
* **participatory design** for developing ideas with affected people;
* **usability testing** for observing task performance.

A persona is useful only when based on evidence and used to focus decisions. A fictional “Maria, age 19, digital native” can hide more than it reveals. Relevant dimensions might include first-language reading, assistive-technology use, research experience, device access, prepaid data, shared-device privacy, and time pressure. Do not turn demographic stereotypes into design requirements.

Recruitment should include people affected by likely barriers. Offering only an inaccessible online sign-up form contradicts inclusive research. Compensation, transport, remote access, language choice, assistive technology, breaks, and consent format may be necessary participation supports.

### Tasks and scenarios

A task is a goal, not a series of interface instructions. Compare:

* weak: “Click Databases, choose A-Z, and search Academic Search.”
* stronger: “You need two recent peer-reviewed articles about coastal community archives for a paper due tomorrow. Find one article you can access and show how you would save its citation.”

The stronger scenario supplies context and success evidence without teaching the interface. A test protocol should define starting state, task, success criteria, maximum time if relevant, permitted assistance, observation fields, and closing questions.

Tasks should represent critical and risky journeys, not only the homepage. Examples include:

* find today's branch hours and confirm an unexpected closure;
* locate a known book and determine where and when it is available;
* discover an article and obtain accessible full text;
* request an archival appointment without exposing sensitive research interests;
* deposit a thesis and understand license and embargo choices;
* recover from a failed login or interrupted form.

### Information architecture and metadata language

**Information architecture** concerns how content and functions are grouped, labelled, related, and navigated. Library interfaces inherit professional structures that may not match public language. `OPAC`, `e-resources`, `serials`, `vertical file`, `finding aid`, and `institutional repository` can be accurate terms but require context or alternatives for novice users.

Metadata choices become interface choices. Facet names, format labels, availability states, and relationship displays determine what users can understand. A facet labelled `Electronic resource` may combine e-books, streamed video, datasets, and web pages. A record labelled `Available` may still require campus authentication. UX work therefore extends into cataloguing, knowledge-base maintenance, and systems configuration.

Card sorting can investigate how users group content. Tree testing can examine whether a proposed hierarchy supports finding. Neither method determines institutional policy by majority vote. Professional, legal, archival, or cultural requirements may justify structures that need explanation rather than removal.

### Plain language and multilingual content

Plain language helps readers find, understand, and use information [@plainlanguage2026]. It involves organization, headings, familiar words, short focused sentences, explicit actions, and useful error messages. It does not mean deleting necessary technical precision.

Compare:

> Authentication unsuccessful.

with:

> We could not sign you in. If you are off campus, start from the library's database page and choose your institution. Your search terms have been kept. Contact the library if the problem continues.

The second message states what happened, suggests recovery, and reassures the user about lost work. It should also expose a technical reference code for support without requiring users to interpret it.

Translation requires more than replacing strings. Navigation labels may expand, date and address formats differ, and professional terms may lack direct equivalents. Interfaces should identify language changes, preserve page language metadata, and avoid mixing untranslated controls into translated content. Local language testing is necessary because an English layout cannot predict every result.

### Accessibility and disability

Web accessibility supports people with visual, auditory, physical, speech, cognitive, language, learning, and neurological disabilities [@w3c2026waiintro]. Disability may be permanent, temporary, or situational. Captions help Deaf users and also people in noisy spaces. Large touch targets help users with limited dexterity and people standing in a moving vehicle. Clear instructions help cognitive access and novice users.

Assistive technologies include screen readers, magnifiers, refreshable braille displays, switch devices, speech input, alternative keyboards, and system-level settings. People also use browser zoom, reader modes, increased text size, high contrast, reduced motion, captions, and keyboard navigation.

Semantic HTML provides structure and behavior to browsers and assistive technologies. Native headings, links, buttons, lists, tables, labels, and landmarks should be used before custom components. ARIA can add accessibility semantics when needed, but incorrect ARIA can make an interface less usable. The WAI-ARIA Authoring Practices Guide documents patterns and keyboard behavior for complex widgets [@w3c2026ariaapg].

### WCAG 2.2 and POUR

WCAG 2.2 organizes guidance under four principles: **perceivable, operable, understandable, and robust** [@w3cwcag22]. Success criteria have conformance levels A, AA, and AAA. Many policies target Level AA, but legal and institutional requirements must be verified locally.

@tbl-pour-library connects selected concerns to LIS services. It is an introduction, not a conformance checklist.

```{r}
#| label: tbl-pour-library
#| tbl-cap: "WCAG principles applied to library services"
#| results: asis
pour_library <- data.frame(
  Principle = c("Perceivable", "Operable", "Understandable", "Robust"),
  Question = c("Can users perceive content through different senses or alternatives?",
               "Can users operate controls with varied input methods and enough time?",
               "Can users understand content, navigation, status, and errors?",
               "Can current and future user agents and assistive technologies interpret the service?"),
  LIS_example = c("Alt text for exhibit images; captions and transcripts; sufficient contrast; reflowed text",
                  "Keyboard-accessible facets; visible focus; skip links; adequate targets; no keyboard trap",
                  "Clear availability labels; consistent navigation; identified language; useful form errors",
                  "Valid semantic elements; programmatic labels, names, roles, values, and status messages"),
  Common_failure = c("Scanned PDF has no text layer or reading order", "Date picker or modal cannot be completed by keyboard",
                     "Session expires without warning and deletes a request", "Custom control looks like a button but exposes no button role")
)
ict_table(pour_library, "WCAG principles applied to library services", "pour-library", c(1, 2.1, 2.3, 2.1))
```

As @tbl-pour-library shows, accessibility spans content, code, and workflow. A perfectly labelled download link still fails if the linked PDF is inaccessible.

Selected checks for an introductory review include:

* meaningful page title and one clear top-level heading;
* logical heading hierarchy;
* meaningful images with appropriate alt text and decorative images ignored;
* text and interface contrast;
* page usable at 200% zoom and narrow reflow without loss;
* every interactive element reachable and operable by keyboard;
* visible focus and logical focus order;
* descriptive link names rather than repeated `click here`;
* explicit form labels, instructions, required states, and error recovery;
* captions and transcripts for media;
* page language identified;
* status messages exposed without unexpectedly moving focus.

W3C provides preliminary “easy checks” and tutorials, while warning that a quick review cannot determine full conformance [@w3c2026easychecks; @w3c2026tutorials].

### Alternative text and complex cultural objects

Alt text communicates an image's purpose in context. The same image may need different treatment in a search result, decorative banner, exhibit, or scholarly analysis. `image.jpg` and “picture of an image” are not useful. A decorative flourish may use empty alt text. A photograph central to an exhibit needs concise identifying information, while a complex map may need a short alternative plus a longer description or accessible data.

Alt text does not replace cataloguing metadata, captions, transcriptions, or scholarly description. A caption identifies and contextualizes an object for all users. Alt text supports the image's function and information when it is not seen. Cultural sensitivity also matters: description should not expose restricted knowledge or treat people as scenery.

### Keyboard access and focus

Keyboard testing is one of the most productive manual checks. Starting at the address bar, use `Tab`, `Shift+Tab`, `Enter`, `Space`, arrow keys where a documented widget pattern requires them, and `Escape` for dismissible overlays. Observe:

* Can every interactive item be reached?
* Is the focus indicator visible?
* Does focus order follow meaning?
* Can menus, filters, modals, and date pickers be operated?
* Does focus become trapped or disappear?
* When a modal closes, does focus return logically?

Do not remove browser focus outlines without providing an accessible replacement. Visual order created with CSS should not contradict keyboard or reading order.

### Accessible forms and authentication

Library forms collect searches, requests, deposits, appointments, and personal data. A placeholder is not a durable label; it disappears when users type and may have low contrast. Native HTML labels and controls provide strong defaults [@whatwg2026forms].

Good forms:

* ask only necessary information;
* group related fields with meaningful legends;
* explain why sensitive information is needed;
* identify required fields in text and programmatically;
* provide input purpose and appropriate autocomplete where safe;
* preserve valid input after errors;
* place error summaries and field errors where users can find them;
* avoid short timeouts or warn and allow extension;
* provide confirmation and correction for consequential submissions.

Authentication deserves end-to-end testing. A library homepage may be accessible while a vendor sign-in, CAPTCHA, multifactor prompt, or proxy error blocks access. Procurement and incident reporting must cover third-party components.

### Responsive and mobile-first services

Responsive design adapts layout and presentation to available space. Mobile-first design begins with the core content and tasks under constraint, then progressively enhances larger screens. It should not mean removing essential services from mobile users.

```{mermaid}
%%| label: fig-responsive-service
%%| fig-cap: "A mobile library task depends on more than viewport width"
flowchart TB
    U["User goal: find and access one resource"] --> D["Device and input"]
    U --> C["Connection, data cost, and interruption"]
    U --> A["Ability, assistive technology, and settings"]
    U --> L["Language, literacy, and domain knowledge"]
    U --> P["Privacy and shared-device context"]
    D --> X["Responsive, resilient service"]
    C --> X
    A --> X
    L --> X
    P --> X
    X --> O["Search, understand, access, recover, and get help"]
```

@fig-responsive-service emphasizes that mobile context is not a screen-size category. A desktop user may have slow connectivity; a phone user may connect a keyboard; a shared device changes privacy needs.

Mobile service design should consider:

* prioritize core tasks and content before institutional promotion;
* use responsive images and avoid autoplay media;
* minimize scripts, fonts, trackers, and unnecessary network requests;
* make touch targets adequately sized and spaced;
* support portrait and landscape orientation unless essential;
* avoid horizontal scrolling for ordinary content;
* allow browser zoom and text resizing;
* keep labels visible and controls close to their purpose;
* preserve entered data across brief interruptions;
* show loading, offline, stale, and retry states clearly;
* provide low-bandwidth alternatives and contact channels.

W3C mobile guidance connects small screens, limited input, bandwidth, and device diversity to web design [@w3c2008mobile]. Mobile accessibility guidance further explains that accessibility requirements apply across mobile web and application contexts [@w3c2015mobile].

### Performance is an access issue

A large page consumes time, data, battery, and attention. Performance barriers disproportionately affect users with older devices, prepaid data, intermittent networks, and assistive technologies processing complex pages. Measure actual transferred bytes, request counts, and loading behavior on throttled connections rather than assuming institutional Wi-Fi.

The fastest component is one not sent. Compress images, use appropriate dimensions, lazy-load noncritical media, limit third-party scripts, cache stable assets, and avoid forcing users to download large PDFs before learning their size and format. Performance optimization should not remove text alternatives, labels, or error messages.

### Testing methods: no single tool is enough

Automated accessibility tools can detect missing labels, some contrast failures, duplicate IDs, and structural issues. They cannot decide whether alt text communicates purpose, heading structure makes sense, an error is understandable, or a workflow is dignified. Manual keyboard, zoom, reflow, content, and assistive-technology testing remain necessary. Disabled users' evaluation adds evidence that standards checks cannot supply.

@tbl-evaluation-methods maps methods to strengths and limits.

```{r}
#| label: tbl-evaluation-methods
#| tbl-cap: "Complementary UX and accessibility evaluation methods"
#| results: asis
evaluation_methods <- data.frame(
  Method = c("Automated accessibility scan", "Manual standards review", "Keyboard and zoom test",
             "Screen-reader test", "Task-based usability test", "Performance and network test", "Analytics and support review"),
  Best_at = c("Rapidly detecting machine-testable patterns", "Interpreting selected success criteria in context",
              "Finding focus, order, reflow, and operation barriers", "Experiencing programmatic structure, names, roles, and announcements",
              "Observing expectations, paths, errors, recovery, and confidence", "Measuring load, transfer, and behavior under constraints",
              "Finding aggregate patterns and recurrent service incidents"),
  Cannot_alone_establish = c("Full WCAG conformance or usable experience", "Representative user success",
                             "All assistive-technology behavior", "Experience of every screen-reader user or disability",
                             "Standards conformance or population prevalence", "Content clarity or task relevance",
                             "Motivation, nonuser exclusion, or causation")
)
ict_table(evaluation_methods, "Complementary UX and accessibility evaluation methods",
          "evaluation-methods", c(1.4, 2.3, 2.2))
```

As @tbl-evaluation-methods demonstrates, triangulation matters. A finding becomes more actionable when observation, standards evidence, and service data point to the same barrier.

### Prioritizing findings

Teams cannot fix everything at once, but prioritization must not reduce accessibility to popularity. A barrier affecting fewer users may completely block a legally or professionally essential service.

Consider:

* **impact:** inconvenience, delay, serious error, or complete block;
* **task criticality:** optional exploration versus required access or submission;
* **reach:** how many people or workflows encounter it;
* **frequency:** how often it occurs;
* **equity:** whether it disproportionately excludes a group;
* **evidence confidence:** strength and sources of evidence;
* **effort and dependency:** complexity, vendor control, and time;
* **risk:** privacy, legal, security, rights, or reputational consequence.

@tbl-severity-scale offers a simple scale. Severity describes user impact, not developer difficulty.

```{r}
#| label: tbl-severity-scale
#| tbl-cap: "A user-impact severity scale for findings"
#| results: asis
severity_scale <- data.frame(
  Severity = c("Critical", "Major", "Moderate", "Minor"),
  User_effect = c("Blocks an essential task with no reasonable accessible alternative",
                  "Causes likely failure, serious error, exclusion, or dependence on assistance",
                  "Creates delay, confusion, repeated effort, or avoidable uncertainty",
                  "Creates friction but task remains understandable and achievable"),
  Example = c("Keyboard user cannot submit an archival appointment request",
              "Screen reader receives unlabeled search fields and cannot distinguish them",
              "Mobile filter resets after rotating or returning from a record",
              "Link wording is vague but surrounding context identifies the destination"),
  Response = c("Stop or provide immediate accessible path; escalate", "Prioritize in current improvement cycle",
               "Schedule with evidence and owner", "Repair when adjacent work occurs; monitor")
)
ict_table(severity_scale, "A user-impact severity scale for findings", "severity-scale", c(.7, 2.2, 2.2, 1.7))
```

As @tbl-severity-scale indicates, an accessibility block is not downgraded because only one participant encountered it. Document the affected task, evidence, relevant criterion, recommended outcome, owner, and verification method.

## Philippine and Southeast Asian examples

### Academic library: off-campus article access

A Philippine university tests article access with students using prepaid mobile data. The discovery page performs adequately, but three vendor redirects, institutional sign-in, and a large PDF create failure. The redesign explains the path, retains citation details, offers an HTML or smaller-file option when available, displays file size, and provides a request link when access fails. The library includes vendor accessibility and performance in renewal review.

### Public library: multilingual branch information

A city public library publishes branch hours, closures, events, and contact details in Filipino, English, and a locally used language. Residents help test labels and directions. Essential information appears as text rather than posters alone. Each poster has an accessible text equivalent, while emergency closures can be reached through a low-bandwidth page and SMS-compatible contact route.

### School library: shared devices and privacy

Students use shared tablets. The catalogue avoids displaying borrowing history on the landing page, provides a clear sign-out action, and does not save searches by default. Text can be resized, controls work with a connected keyboard, and reading-list pages avoid motion and distracting autoplay. Teachers and students with disabilities participate in testing.

### Government information service: accessible forms

A government records request form uses visible labels, clear purpose statements, appropriate input types, an error summary, preserved entries, and a confirmation page. Required identity information is explained and minimized. A downloadable form is not offered only as an inaccessible scanned PDF; an assisted offline channel remains available.

### Regional cultural heritage exhibit

A Southeast Asian digital exhibit contains multilingual object labels, transcripts for oral histories, captions for video, alt text for meaningful images, and long descriptions for maps. Community advisors determine whether certain images or knowledge should have restricted access. Responsive image sizes reduce data cost without reducing descriptive access.

> **Reflect:** Think of the last library or government service you used on a phone. Which barrier came from interface design, which from organizational policy, and which from an external provider?

## Hands-on activity: Conduct a mobile UX and accessibility critique

### Purpose

Evaluate one meaningful library task on desktop and mobile using task observation, manual accessibility checks, and a limited automated scan. Produce prioritized findings and a redesign sketch. This activity is the UX or discovery critique worth 10% of the course assessment.

### Tools

* one public library, academic library, school library, repository, archive, museum, or government information website;
* desktop and mobile-sized browser views or two devices;
* keyboard only for one test pass;
* browser zoom and text-resize controls;
* one automated accessibility checker approved by the instructor;
* performance information from browser developer tools when available;
* observation sheet based on @tbl-test-log.

```{r}
#| label: tbl-test-log
#| tbl-cap: "Fields for the mobile UX and accessibility observation log"
#| results: asis
test_log <- data.frame(
  Field = c("Task and success evidence", "Participant and context", "Device and viewport", "Connection context",
            "Observed step", "Expectation or comment", "Barrier and recovery", "Accessibility evidence",
            "Severity", "Recommendation", "Owner and verification"),
  Record = c("Goal, starting point, completion evidence", "Relevant experience only; avoid unnecessary identity data",
             "Device class, input, orientation, zoom", "Normal or throttled; do not collect network identifiers",
             "Action and interface response", "Quote or concise paraphrase with consent",
             "What prevented progress and whether recovery succeeded", "Criterion, keyboard behavior, semantics, contrast, or tool evidence",
             "Critical, major, moderate, or minor with rationale", "Desired outcome, not merely a preferred visual style",
             "Responsible team and retest method")
)
ict_table(test_log, "Fields for the mobile UX and accessibility observation log", "test-log", c(1.2, 3))
```

@tbl-test-log keeps observations separate from interpretations and recommendations.

### Steps

1. Select a service and one critical task, such as finding an available item, accessing an article, requesting archival material, or locating branch information.
2. Write a realistic scenario and observable success criteria. Do not include click-by-click instructions.
3. Define the test scope, participant consent, recording method, data retention, and stopping rule. Do not test a sensitive personal topic.
4. Conduct a pilot yourself. Correct unclear instructions before involving a peer.
5. Ask one peer participant to attempt the task on a mobile viewport while thinking aloud. Remind the participant that the service is being tested, not the person.
6. Observe without teaching. If help is requested, record when and what was provided.
7. Repeat the task independently on a desktop viewport and compare content, order, labels, and functionality.
8. Test the essential path by keyboard. Record reachability, visible focus, focus order, overlays, menus, filters, forms, and recovery.
9. Review at 200% zoom and at a narrow viewport. Check reflow, text size, orientation, horizontal scrolling, and hidden content.
10. Inspect page title, headings, link names, form labels, meaningful images, media alternatives, language, errors, and status messages.
11. Run one automated checker. Verify each reported issue manually and state what the tool did not test.
12. Inspect page transfer size or loading behavior on a throttled connection when tools permit. Record large blockers and failure states.
13. Group observations into findings. Assign severity using user effect, task criticality, reach, equity, and evidence.
14. Create one annotated redesign sketch for the highest-priority finding. Include keyboard, screen-reader, mobile, error, and low-bandwidth behavior as relevant.
15. Write a concise recommendation identifying owner, dependency, expected outcome, and verification method.

### Expected output

Submit:

* test protocol and consent script;
* participant and evaluator observation notes with unnecessary personal data removed;
* annotated desktop and mobile screens;
* keyboard, zoom, reflow, content, and automated-check results;
* five to eight findings ranked by severity;
* one redesigned task flow or wireframe;
* an 800-word evidence-based critique; and
* a 150-word reflection: **Which barrier would professional staff have missed without observing a user or testing a different mode of access?**

### Lab checklist

- [ ] The task represents a meaningful service goal.
- [ ] Success criteria are observable and not tied to one path.
- [ ] Consent and data minimization are documented.
- [ ] Mobile and desktop experiences are compared.
- [ ] Keyboard, zoom, reflow, and content checks are manual.
- [ ] Automated results are verified and not treated as full conformance.
- [ ] Findings distinguish evidence, interpretation, and recommendation.
- [ ] Severity is based on user and task impact.
- [ ] Redesign notes include accessible behavior, not only appearance.
- [ ] Each priority finding has an owner and retest method.

### Assessment rubric

@tbl-ux-lab-rubric rewards evidence, inclusion, and actionable analysis.

```{r}
#| label: tbl-ux-lab-rubric
#| tbl-cap: "Rubric for the mobile UX and accessibility critique"
#| results: asis
ux_lab_rubric <- data.frame(
  Criterion = c("Task and protocol", "Observation and evidence", "Accessibility and mobile analysis",
                "Prioritization and redesign", "Ethics and communication"),
  Weight = c("15%", "25%", "25%", "20%", "15%"),
  Strong_performance = c("Scenario, success criteria, scope, and method align with a real service need",
                         "Notes separate observed behavior from interpretation and triangulate evidence",
                         "Manual and automated methods address POUR, keyboard, reflow, content, forms, and connection constraints",
                         "Severity reflects impact; redesign addresses cause and includes a verification plan",
                         "Consent, privacy, respectful facilitation, limitations, and accessible reporting are clear")
)
ict_table(ux_lab_rubric, "Rubric for the mobile UX and accessibility critique",
          "ux-lab-rubric", c(1.2, .6, 3))
```

## Ethical and professional issue: surveillance, dark patterns, and digital exclusion

UX research and analytics can improve services, but they can also become surveillance. Search queries, click paths, account events, device details, locations, recordings, and session replays may reveal health, political, religious, legal, academic, or identity-related interests. Collecting data because a tool offers it is not a legitimate research purpose.

Ethical UX research requires a specific question, informed consent where appropriate, minimum necessary data, secure storage, limited access, a retention schedule, and reporting that avoids re-identification. Participants must know they may stop. Researchers should not record passwords, patron records, or unrelated screen content. The Philippine Data Privacy Act's principles of transparency, legitimate purpose, and proportionality apply to research and analytics [@philippines2012dpa; @npc2023dpa].

**Dark patterns** are interface choices that manipulate people toward actions that benefit the service provider rather than the user. Examples include making privacy-protective options difficult to find, using confusing consent language, preselecting broad data sharing, disguising advertising as search results, or making account deletion harder than account creation. Libraries should not reproduce these practices through vendor defaults.

Digital exclusion also occurs when a service assumes current smartphones, stable broadband, individual email accounts, English fluency, credit cards, or comfort with identity verification. Offering an “accessible alternative” through a slower, stigmatizing, or staff-dependent route may preserve formal access while denying equal independence.

Professional safeguards include:

* include disabled and marginalized users throughout planning and evaluation;
* provide accessible participation and fair compensation;
* minimize analytics and disable invasive replay by default;
* separate service improvement data from patron profiling;
* retain raw recordings only as long as necessary;
* explain personalization and allow nonpersonalized use;
* require accessibility, privacy, performance, and exit clauses in procurement;
* publish accessible contact, correction, and complaint channels;
* maintain offline or assisted routes without making them inferior;
* report limitations and populations not represented in a study.

> **Professional judgment prompt:** A vendor offers session recordings showing every tap, keystroke, and form entry. What research question could justify any part of this collection, which fields must be masked or excluded, who may view recordings, and when must they be deleted?

## Review questions

1. **Identification:** Distinguish UX, usability, accessibility, inclusive design, and responsive design.
2. **Identification:** What are WCAG's four POUR principles?
3. **Short answer:** Why can neither an automated scan nor one usability test establish that a service is accessible?
4. **Short answer:** Explain how metadata labels and availability statements affect UX.
5. **Applied design:** Rewrite “Click Advanced Search and select Format: Serial” as a realistic task scenario.
6. **Applied accessibility:** A filter opens with a mouse but cannot receive keyboard focus. Identify the user effect, likely WCAG principle, severity, and retest.
7. **Scenario:** A mobile form clears all entries after a brief connection loss. Recommend content, technical, and recovery changes.
8. **Scenario:** A library's translated site retains English error messages and unlabeled icons. Why is translation alone insufficient?
9. **Scenario:** Analytics show few screen-reader users. Explain why this is not evidence that accessibility work has low priority.
10. **Critical analysis:** When might simplifying a service for mobile become harmful removal rather than useful prioritization?

## Mini case: The one-box redesign

Sampaguita University launches a new discovery homepage built around one large search box. Senior managers praise its clean appearance. Mobile analytics show fewer searches ending in full-text access, but the vendor dashboard records only page views and clicks. Student testing reveals that filters are hidden after zoom, the sign-in prompt appears only after users select an article, and failed authentication deletes the query. A blind student reports repeated unlabeled “View” links. A student using prepaid data abandons pages with large promotional images. Librarians suggest adding instructions everywhere; the vendor proposes collecting session recordings to learn more.

### Guide questions

1. Which usability, accessibility, mobile, metadata, authentication, performance, analytics, and privacy problems are present?
2. What mixed-method evaluation would produce useful evidence without collecting passwords, queries, or unnecessary personal data?
3. Prioritize three interventions, identify responsible owners, and specify how each would be verified with affected users.

## Further reading and tools

### Readings and standards

* **Web Content Accessibility Guidelines 2.2.** The central W3C accessibility recommendation used in this chapter [@w3cwcag22].
* **Introduction to Web Accessibility.** A concise W3C overview of disability, barriers, and benefits [@w3c2026waiintro].
* **Easy Checks.** A guided preliminary review that explicitly does not establish full conformance [@w3c2026easychecks].
* **WAI Tutorials.** Practical guidance on page structure, menus, images, tables, forms, and carousels [@w3c2026tutorials].
* **ARIA Authoring Practices Guide.** Patterns and keyboard interaction for complex web widgets [@w3c2026ariaapg].
* **Mobile accessibility guidance.** W3C discussion connecting accessibility guidelines to mobile contexts [@w3c2015mobile].
* **Library usability studies.** Examples of applying usability and accessibility inquiry in academic library websites [@billingham2014accessibility; @emde2009usability; @vargas2026homepage].

### Open or widely accessible tools

* **Browser keyboard, zoom, responsive view, and developer tools:** essential manual testing capabilities.
* **WAVE, axe DevTools, Accessibility Insights, or Lighthouse:** automated support for finding selected issues; none proves conformance.
* **NVDA:** free Windows screen reader. Testing requires learning expected screen-reader interaction rather than treating speech output as a simple scanner.
* **Colour Contrast Analyser:** desktop tool for testing text and graphical contrast.
* **WebPageTest or browser network throttling:** tools for examining performance under constrained connections; do not submit private URLs or data to public services.
* **Penpot, diagrams.net, paper, or HTML prototypes:** accessible ways to communicate redesigns when accompanied by behavior notes and text alternatives.

### Closing synthesis

UX connects institutional systems to lived experience. Accessibility asks whether people with disabilities can participate independently and effectively. Mobile design exposes assumptions about devices, bandwidth, privacy, interruption, and attention. These are not separate finishing tasks: they shape metadata labels, content, authentication, vendor selection, help, assessment, and policy.

An attractive interface can still exclude. A standards-conformant page can still confuse. A fast mobile page can still manipulate consent. LIS professionals therefore need multiple forms of evidence and sustained participation by affected users. The governing question is: **Who can complete which information task, under what conditions, with what effort and dignity, and who is missing from the evidence?** Chapter 10 extends this lifecycle responsibility from immediate access to long-term digital continuity, preservation, storage, and cloud services.
