---
title: "Networking, Web Basics, and Library Websites"
---

```{r}
#| label: setup
#| include: false

latex_escape <- function(x) {
  x <- gsub("\\", "<<<BACKSLASH>>>", as.character(x), fixed = TRUE)
  x <- gsub("([#$%&_{}])", "\\\\\\1", x, perl = TRUE)
  x <- gsub("~", "\\textasciitilde{}", x, fixed = TRUE)
  gsub("<<<BACKSLASH>>>", "\\textbackslash{}", x, fixed = TRUE)
}

ict_table <- function(data, caption, label = NULL, widths = NULL) {
  data <- as.data.frame(data, stringsAsFactors = FALSE)
  names(data) <- gsub("_", " ", names(data), fixed = TRUE)
  anchor <- if (!is.null(label)) paste0("{#tbl-", label, "}") else ""
  if (knitr::is_latex_output()) {
    align <- paste(rep("Y", ncol(data)), collapse = "")
    header <- paste(vapply(names(data), latex_escape, character(1)), collapse = " & ")
    rows <- apply(data, 1, function(row) paste(vapply(row, latex_escape, character(1)), collapse = " & "))
    out <- c(
      sprintf("\\begin{table}[htbp]\\centering\\caption{%s}\\label{tbl-%s}", latex_escape(caption), label),
      sprintf("\\begin{tabularx}{\\textwidth}{%s}", align),
      "\\toprule", paste0(header, " \\\\"), "\\midrule",
      paste0(rows, " \\\\"), "\\bottomrule", "\\end{tabularx}", "\\end{table}"
    )
  } else if (knitr::is_html_output()) {
    out <- knitr::kable(data, format = "html", caption = caption, escape = TRUE,
                        table.attr = if (is.null(label)) "" else sprintf("id='tbl-%s'", label))
  } else {
    out <- c(paste0(": ", caption, " ", anchor), "", knitr::kable(data, format = "pipe"))
  }
  knitr::asis_output(paste(out, collapse = "\n"))
}
```

## Chapter overview

A library website may be open twenty-four hours a day, but that does not mean every user can reach or use it. A student taps a database link on a phone. The page takes too long to load, redirects to an unfamiliar login screen, and reports that the session has expired. A parent searches a public-library site for today's opening hours but finds an outdated poster. A researcher follows a repository link from a search engine and receives a "404 Not Found" response. These are not merely website inconveniences. They are failures in the delivery of information services.

The central problem of this chapter is **how to make a library service reliably reachable and understandable across networks, browsers, devices, and user contexts**. Solving that problem requires more than visual design. It requires a working model of networks, internet addressing, the Domain Name System (DNS), URLs, HTTP and HTTPS, servers, browsers, HTML, CSS, content management, accessibility, performance, and website governance.

Chapter 1 described an information service as a layered sociotechnical system. This chapter opens two of those layers: network infrastructure and the web interface. You will trace a request from a browser to a library server and back, then create a small semantic web page. The aim is not to train you as a network engineer or professional web developer. It is to give you enough technical understanding to diagnose service problems, communicate requirements, maintain trustworthy content, and collaborate effectively with IT staff, vendors, designers, and users.

## Learning objectives

After completing this chapter, you should be able to:

1. **Distinguish** the internet, the World Wide Web, local networks, and web-based library services.
2. **Diagram** a client-server exchange involving a device, network, DNS, server, HTTP request, and response.
3. **Interpret** the principal parts of a URL and common HTTP status codes.
4. **Explain** how HTML, CSS, and JavaScript contribute different functions to a web page.
5. **Create** a small semantic HTML page with headings, links, lists, contact information, and basic accessibility features.
6. **Evaluate** a library website for findability, security, accessibility, mobile use, low-bandwidth performance, and governance.

## Key concepts

@tbl-networking-key-concepts introduces the vocabulary used throughout the chapter. Several concepts operate at different layers. DNS helps a client locate a server, while HTML gives structure and meaning to the document returned by that server.

```{r}
#| label: tbl-networking-key-concepts
#| results: asis
networking_key_concepts <- data.frame(
  Term = c("Network", "LAN", "WAN", "Internet", "World Wide Web", "Protocol",
           "Packet", "Bandwidth", "Latency", "Client", "Server", "IP address",
           "Domain name", "DNS", "URL", "HTTP", "HTTPS", "Browser", "HTML",
           "CSS", "JavaScript", "Hosting", "CMS", "Responsive design",
           "Semantic HTML"),
  Working_definition = c(
    "Connected devices that exchange data and share resources.",
    "A local area network covering a limited site such as a building or campus.",
    "A wide area network connecting sites across larger geographic areas.",
    "The global system of interconnected networks using the Internet Protocol suite.",
    "Interlinked resources identified by URLs and transferred mainly through HTTP or HTTPS.",
    "An agreed set of rules for communication between systems.",
    "A formatted unit into which data is divided for network transmission.",
    "The amount of data a connection can carry during a period of time.",
    "The delay between sending a request and receiving a response.",
    "A device or program that requests a service or resource.",
    "A device or program that responds to requests and provides resources.",
    "A numerical address used to identify and route traffic to a network interface.",
    "A human-readable name associated with internet resources.",
    "The distributed system that translates domain names into network information.",
    "A structured address identifying a resource and how to access it.",
    "The application protocol used to request and transfer web representations.",
    "HTTP protected by Transport Layer Security.",
    "Software that requests, interprets, and presents web resources.",
    "The markup language representing the structure and meaning of web content.",
    "A style language controlling presentation and responsive layout.",
    "A programming language adding browser behavior and dynamic interaction.",
    "Server, storage, network, and related resources that make a site available.",
    "A content management system for creating, organizing, and publishing content.",
    "Layouts and content that adapt to different screens and devices.",
    "HTML elements used according to meaning and purpose rather than appearance alone."
  )
)
ict_table(networking_key_concepts, "Key concepts for networking and library websites",
          "networking-key-concepts", c(1, 3))
```

## Core discussion

### Networks as service infrastructure

A **network** exists when devices exchange data according to agreed rules. In a small library, a local area network may connect staff computers, printers, self-check stations, wireless access points, an on-site server, and a router. A university may connect several campus libraries through a larger institutional network. Remote branches, cloud systems, and vendor platforms depend on wide area and internet connections.

Networks allow information organizations to share catalogues, storage, authentication, printers, internet access, and applications. They also create dependencies. If the local network fails, a cloud-hosted circulation system may be unreachable even though the vendor server operates normally. If the internet connection works but DNS fails, a service may be reachable by numerical address while its familiar name appears unavailable. Diagnosis begins by locating the layer at which communication stopped.

Network data are divided into **packets**, transmitted through links and routing devices, and reassembled at the destination. Two measures help explain user experience:

* **Bandwidth** concerns how much data a connection can carry. Large images and video load slowly on a low-bandwidth link.
* **Latency** concerns delay. Even a small page can feel slow when it requires many round trips over a high-latency connection.

Bandwidth and latency are not interchangeable. Website teams should measure whether users can complete real tasks rather than treating one advertised connection speed as a full description of service quality.

### Internet and web are not synonyms

The **internet** is the global network infrastructure. The **World Wide Web** is one service that uses it. Email, file transfer, voice communication, time synchronization, and many APIs also use the internet but are not identical to the web.

The distinction matters. A library can have a working internet connection while its website is unavailable. A catalogue hosted inside a building may work on the LAN during an internet outage but be unreachable from home. An offline exhibit installed on a kiosk can use web technologies without being published on the internet. Accurate language helps teams identify the actual service boundary.

### Protocols and layers

A **protocol** is an agreed set of communication rules. At a simplified level, IP supports addressing and routing; TCP supports a reliable ordered connection for many applications; DNS associates human-readable names with network information; HTTP defines web requests and responses; and TLS protects data in transit for HTTPS.

Students do not need to memorize the full protocol stack. The useful habit is to ask which protocol is responsible for which task. A DNS problem, an expired TLS certificate, and an HTTP "Not Found" response can all block access, but they require different remedies and responsible staff.

### From a URL to a page

@fig-web-request traces a simplified request for a library-hours page. The browser interprets the URL, asks DNS for an address, connects through network infrastructure, establishes a protected session for HTTPS, sends an HTTP request, receives a response, and interprets the returned HTML and related resources.

```{mermaid}
%%| label: fig-web-request
%%| fig-cap: "A simplified request path from a library user to a web service"
sequenceDiagram
    participant U as User
    participant B as Browser client
    participant D as DNS resolver
    participant N as Networks
    participant S as Library web server
    U->>B: Enter or follow an HTTPS URL
    B->>D: Request address for the domain
    D-->>B: Return network address
    B->>N: Connect and establish protected session
    N->>S: Deliver HTTP request
    S-->>N: Return status, headers, and representation
    N-->>B: Deliver HTTP response
    B->>B: Interpret HTML, CSS, and other resources
    B-->>U: Present the page and actions
```

@fig-web-request reveals that the browser is an active client. It sends headers, manages cache entries and cookies, validates secure connections, constructs a document model, applies styles, and may run scripts. Browser differences, extensions, privacy settings, and older devices can therefore affect service behavior.

### IP addresses, domain names, and DNS

An **IP address** identifies a network interface for communication and routing. People remember names more easily than numerical addresses, so services use names such as library.example.edu.

DNS is a distributed hierarchical naming system rather than one universal address book. Its foundational design separates the name space from the many servers that answer queries about portions of it [@mockapetris1987dns]. DNS information is cached for a defined period, so changes may not appear everywhere immediately.

Domain governance is an LIS concern because names support continuity, trust, and citation. An institution should know who controls registration, who manages DNS records, when registrations and certificates expire, how changes are authorized, and which services depend on each name. A public service should not depend permanently on a domain registered under one employee's private account.

### Reading a URL

Consider:

```text
https://library.example.edu/research/guides.html?subject=history#databases
```

Its scheme is https; its host is library.example.edu; its path is /research/guides.html; its query is subject=history; and its fragment is databases. URLs appear in catalogues, citations, QR codes, syllabi, and archival records. A stable URL policy therefore supports discovery and continuity.

Changing a URL without a redirect breaks inbound links. A migration should inventory important URLs, map old addresses to new ones, configure permanent redirects, and monitor "Not Found" reports. Human-readable paths help users judge purpose, but appearance alone is not proof of safety. Users should inspect the complete host name before entering credentials.

### HTTP requests and responses

HTTP is a request-response protocol. RFC 9110 defines current HTTP semantics for methods, status codes, fields, and representations [@fielding2022http]. A request includes a method, target, headers, and sometimes a body. A response includes a status code, headers, and often HTML, JSON, an image, or a PDF.

Common methods include GET for retrieving a representation, HEAD for retrieving response metadata, POST for submitting data, and PUT or PATCH for creating or modifying resources in systems that support them.

@tbl-http-status lists status codes that LIS staff may encounter. A status is evidence, not a complete diagnosis. A 404 response says that the server found no current resource at that path; it does not explain whether the resource was removed, renamed, mistyped, or migrated without a redirect.

```{r}
#| label: tbl-http-status
#| results: asis
http_status <- data.frame(
  Code = c("200", "301", "302", "400", "401", "403", "404", "429", "500", "503"),
  Meaning = c("OK", "Moved Permanently", "Temporary redirect", "Bad Request",
              "Authentication required", "Forbidden", "Not Found",
              "Too Many Requests", "Internal Server Error", "Service Unavailable"),
  LIS_interpretation = c(
    "The requested representation was returned.",
    "Use the new stable location; old links should normally redirect.",
    "The client is being sent temporarily elsewhere.",
    "Inspect the URL, form, or submitted data.",
    "The user must authenticate or supplied credentials failed.",
    "The server understood the request but refuses access.",
    "Check deletion, migration, spelling, and link maintenance.",
    "Relevant to APIs, crawlers, and automated harvesting.",
    "The application failed while processing the request.",
    "The service is temporarily unavailable because of maintenance or overload."
  )
)
ict_table(http_status, "Common HTTP status codes in information services",
          "http-status", c(.5, 1.2, 2.5))
```

### HTTPS, certificates, and trust

HTTPS protects HTTP traffic with TLS. Correct configuration supports confidentiality, integrity in transit, and server authentication. It does not prove that content is accurate or ethical, and it does not protect data after they reach an insecure server. It secures an important channel, not the entire service.

Certificate warnings may indicate expiration, a host-name mismatch, incomplete configuration, or interception. Libraries should monitor certificates and renew them before expiration. Staff should never make bypassing a warning the routine solution.

### HTML, CSS, and JavaScript

The browser commonly receives HTML, then requests style sheets, scripts, images, and fonts. @fig-web-page-layers separates three technologies that beginners often confuse.

```{mermaid}
%%| label: fig-web-page-layers
%%| fig-cap: "The complementary roles of HTML, CSS, and JavaScript"
flowchart LR
    H["HTML<br/>structure and meaning"] --> P["Usable web page"]
    C["CSS<br/>presentation and responsive layout"] --> P
    J["JavaScript<br/>behavior and interaction"] --> P
    S["Content, metadata, links, and policy"] --> H
    A["Accessibility and progressive enhancement"] --- H
    A --- C
    A --- J
```

HTML identifies headings, paragraphs, navigation, lists, forms, tables, and images. CSS controls presentation and responsive layout. JavaScript adds behavior such as menu expansion, form validation, and asynchronous updates. The WHATWG standard defines HTML as a language for structuring web documents and applications [@whatwg2026html].

As @fig-web-page-layers shows, technology does not replace content governance. A large bold paragraph may look like a heading but is not programmatically identified as one. A clickable generic container styled like a button lacks the native semantics and behavior of a real button. Semantic HTML helps browsers, screen readers, search engines, and maintenance tools interpret meaning.

### A minimal semantic library page

This example publishes an essential service with meaningful structure:

```html
<!doctype html>
<html lang="en">
<head>
  <meta charset="utf-8">
  <meta name="viewport" content="width=device-width, initial-scale=1">
  <title>Library Hours | Mabini Community Library</title>
</head>
<body>
  <header>
    <h1>Mabini Community Library</h1>
    <nav aria-label="Primary">
      <a href="index.html">Home</a>
      <a href="catalogue.html">Catalogue</a>
      <a href="contact.html">Contact</a>
    </nav>
  </header>
  <main>
    <h2>Library hours</h2>
    <p>Monday-Friday: 8:00 a.m.-5:00 p.m.</p>
    <p>Saturday: 9:00 a.m.-12:00 noon</p>
    <h2>Before you visit</h2>
    <ul>
      <li>Bring an ID for registration.</li>
      <li>Contact us before requesting archival materials.</li>
    </ul>
  </main>
  <footer>
    <address>Email: <a href="mailto:library@example.org">library@example.org</a></address>
    <p>Last reviewed: 18 June 2026</p>
  </footer>
</body>
</html>
```

The language attribute supports language-aware tools. The character encoding supports consistent display. The viewport instruction supports mobile layouts. The title distinguishes the page in browser tabs and search results. Headings communicate organization, link text names destinations, and the review date supports content governance.

### Progressive enhancement and resilience

**Progressive enhancement** begins with meaningful content and basic functions that work under varied conditions, then adds richer presentation and behavior. Opening hours should remain readable if a script fails. A catalogue destination should be a real link. A form should retain understandable labels when visual styling is unavailable.

This approach does not prohibit modern applications. It asks teams to identify the essential service and protect it from avoidable dependencies. Resilience matters for older devices, assistive technologies, intermittent signals, data-saving settings, and restricted institutional computers.

### Library websites as service gateways

A library website is not an online brochure. It connects users to catalogues, databases, repositories, guides, reference help, policies, events, spaces, accounts, and staff expertise. Its information architecture should reflect user tasks rather than the institution's internal chart.

Common tasks include checking hours, finding a known item, searching a topic, accessing electronic resources, renewing materials, asking for help, learning access rules, and verifying whether information is current and official.

Library usability research demonstrates the value of observing users rather than assuming professional terms are self-explanatory [@emde2009usability]. Accessibility research similarly shows that meaningful access requires continuing attention rather than a one-time launch check [@billingham2014accessibility]. Recent homepage work still connects structural and visual decisions with usability and accessibility [@vargas2026homepage].

### Content management and hosting

A **content management system** may provide templates, navigation, revision histories, user roles, scheduled publication, and approval workflows. It can improve distributed publishing, but it creates governance and security work. Plugins require updates, accounts require review, themes may introduce accessibility problems, and old pages may remain searchable. The institution still needs content owners, review schedules, backup procedures, and an exit plan.

Hosting may use institutional servers, a managed provider, cloud infrastructure, or a hosted software service. The choice affects control, expertise, security updates, cost, performance, jurisdiction, backup, export, and vendor dependence. "In the cloud" does not remove institutional responsibility.

Remote access may involve a proxy, VPN, or federated identity service. Each transition should tell users where they are going and why authentication is required. Repeated redirects and unclear domains can make legitimate access resemble phishing. Libraries should document operators, data exchange, support responsibility, and what happens when a subscription ends.

### Accessibility, mobile use, and performance

WCAG 2.2 requires web content to be perceivable, operable, understandable, and robust [@w3cwcag22]. Mobile accessibility is not separate: small screens, touch input, orientation, magnification, and device constraints interact with the same principles [@w3c2015mobile].

Beginner checks include meaningful titles, logical headings, text alternatives, form labels, keyboard operation, adequate contrast, text reflow, clear link purpose, and understandable errors. Accessibility benefits many users, but that broad benefit should not erase the specific rights and expertise of persons with disabilities.

Performance is also an access issue. Large posters, multiple fonts, trackers, and third-party widgets consume data and add failure points. Regional digital plans recognize that meaningful participation depends on both infrastructure and capability [@asean2021digital]. Practical measures include compressing images, avoiding autoplay, limiting external scripts, offering text alternatives, making essential information available without accounts, and testing on a low-cost phone and constrained connection.

### Website governance

Library websites are published information resources and organizational records. @tbl-web-governance connects components to accountable questions. No component is purely technical: each has a service owner, policy implication, and maintenance need.

```{r}
#| label: tbl-web-governance
#| results: asis
web_governance <- data.frame(
  Component = c("Domain and DNS", "Hosting and platform", "Page content",
                "Links and redirects", "Accounts and permissions",
                "Analytics and logs", "Backups and archives"),
  Accountable_question = c(
    "Who controls registration, records, renewal, and emergency changes?",
    "Who patches, monitors, backs up, and supports the service?",
    "Who owns accuracy, accessibility, approval, and review dates?",
    "Who tests destinations and preserves high-value old URLs?",
    "Who receives privileges, and when are they reviewed or removed?",
    "What is collected, why, for how long, and with whose access?",
    "What can be restored or preserved, and has recovery been tested?"
  ),
  Evidence = c("Domain inventory and renewal record", "Agreement and maintenance log",
               "Content register and review history", "Link report and redirect map",
               "Role list and access review", "Data inventory and retention rule",
               "Backup log, archive policy, and recovery test")
)
ict_table(web_governance, "Website governance responsibilities and evidence",
          "web-governance", c(1, 2.3, 1.5))
```

As @tbl-web-governance suggests, publishing is a lifecycle. A page needs a purpose, audience, owner, approval status, review date, and eventual disposition. Event notices expire quickly; policy pages need revision histories; research guides require subject review; and significant exhibits may warrant preservation.

> **Reflective prompt.** Recall a website that prevented you from completing a task. Which layer failed: connection, naming, security, response, page structure, content, accessibility, or governance? What evidence supports your diagnosis?

## Philippine and Southeast Asian examples

### State university library: the remote-access chain

A Philippine state university links subscribed databases from its library website. On campus, access is recognized by network address. Off campus, students authenticate through a proxy. Success depends on current vendor links, DNS, proxy configuration, identity records, HTTPS certificates, license rules, and help documentation. When a vendor changes its domain, old proxy links fail. The library responds with a link inventory, off-campus testing, clear escalation information, and coordinated updates.

### Municipal library: hours during a typhoon

A municipal library normally posts hours as a colorful image. During a typhoon, staff update social media but forget the website image and PDF. Residents see three schedules. The library redesigns the page so current status appears as lightweight HTML with an update time. One role approves emergency updates, and expired notices are removed through a documented workflow.

### School library: service during internet outages

A rural school library uses a local server to provide reading materials over campus Wi-Fi. The collection remains available inside the school when the external connection fails. Staff distinguish the local web from the internet, maintain offline backups, and use a text-first interface suitable for older phones.

### Regional heritage portal: multilingual stable links

A Southeast Asian heritage consortium aggregates collections from several museums. Partners use different scripts, languages, domains, and systems. They adopt Unicode, language metadata, institutional identifiers, persistent item URLs, and redirects. The interface presents multilingual values and provenance rather than replacing local description with one regional label.

### Public information service: external maps and privacy

A public library embeds a commercial map and analytics on every page. The map helps some visitors but adds tracking and a failure dependency. The library provides its address, landmarks, transport directions, and telephone number as HTML before the map. It then evaluates whether analytics can collect less data.

## Hands-on activity: Trace and build a library web page

### Purpose

Trace a real web request, then create a semantic HTML page that communicates an essential library service under mobile and low-bandwidth conditions.

### Tools

* A browser with developer tools, such as Firefox, Chrome, or Edge
* A text editor such as Notepad, Visual Studio Code, or Positron
* The W3C Nu HTML Checker at <https://validator.w3.org/nu/>
* A folder named **chapter2-lab**

### Steps

1. Open a library, archive, museum, or repository page that requires no personal information.
2. Record its URL and label the scheme, host, path, query, and fragment where present.
3. Open the browser Network panel, reload, and select the main document request.
4. Record the method, status, content type, transferred size, and domain.
5. Count contacted domains and distinguish institutional from external services.
6. Use a narrow mobile viewport to find hours, the catalogue, or contact help.
7. Create **hours.html** with a document type, language, encoding, viewport metadata, and unique title.
8. Add header, navigation, main, and footer landmarks with one h1 and logical h2 headings.
9. Add fictional hours, a status message, visit requirements, contacts, and a review date.
10. Add three meaningful links. If using an image, resize it and supply useful alternative text.
11. Test two viewport widths, validate the HTML, and correct errors you understand.
12. Ask a peer to locate Saturday hours and a contact method without explanation.

### Expected output

Submit the annotated request record, three network and mobile observations, **hours.html**, validation evidence, the peer-task result, and a 150-word reflection.

> **Reflection question.** Which improvement came from understanding the network, which came from HTML structure, and which came from observing a user?

### Activity checklist

@tbl-ch2-lab-checklist moves from technical validity to service validity. The page must load and be structured correctly, but it must also communicate current, usable information.

```{r}
#| label: tbl-ch2-lab-checklist
#| results: asis
ch2_checklist <- data.frame(
  Check = c("URL parts, method, status, type, size, and domain are recorded",
            "External domains and one dependency risk are identified",
            "Document includes language, encoding, viewport, and unique title",
            "Semantic landmarks and logical headings are used",
            "Essential hours and contact information are HTML text",
            "Links have meaningful text and work locally",
            "Page works at narrow width and without its optional image",
            "Validation and peer-task results are documented",
            "No real personal or confidential information is included"),
  Complete = rep("[ ]", 9)
)
ict_table(ch2_checklist, "Chapter 2 web lab submission checklist",
          "ch2-lab-checklist", c(3, .5))
```

### Short lab rubric

@tbl-ch2-lab-rubric allocates 20 points. Visual sophistication is not a scoring category; clarity, semantics, resilience, and evidence matter more.

```{r}
#| label: tbl-ch2-lab-rubric
#| results: asis
ch2_rubric <- data.frame(
  Criterion = c("Request trace", "Semantic HTML", "Access and resilience",
                "Testing and correction", "Reflection and presentation"),
  Points = c(4, 6, 4, 3, 3),
  Full_credit = c(
    "Records request evidence accurately and identifies a dependency",
    "Uses valid metadata, landmarks, headings, lists, links, and contact structure",
    "Essential content works at narrow width without images or complex scripts",
    "Reports validation and peer testing and makes evidence-based corrections",
    "Explains learning across network, document, and user layers"
  )
)
ict_table(ch2_rubric, "Rubric for the Chapter 2 web lab",
          "ch2-lab-rubric", c(1, .4, 2.8))
```

## Ethical and professional issue: access, surveillance, and secure delivery

Library websites can broaden access while excluding, observing, or endangering users. An online-only form may disadvantage people without stable connectivity, compatible devices, data allowance, accessible design, or required identification. The IFLA-UNESCO Public Library Manifesto connects digital access with inclusion and equitable service [@ifla2022manifesto]. Digital channels should expand options rather than silently eliminate necessary alternatives.

Web requests can disclose IP addresses, device information, referring pages, cookies, and behavioral patterns. Third-party fonts, maps, videos, chat widgets, and social-media embeds may transmit data before intentional interaction. Before adding one, ask what is sent, to whom, for what purpose, for how long, and whether the essential service works without it.

The Philippine Data Privacy Act's principles of transparency, legitimate purpose, and proportionality apply to forms, logs, analytics, and vendor integrations [@philippines2012dpa; @npc2023dpa]. A notice cannot by itself legitimize unnecessary collection.

HTTPS should be standard, especially for authentication, forms, and accounts. Interface choices also influence security. A database link that silently opens an unfamiliar domain and asks for credentials may train users to ignore phishing indicators. The page should name the provider or authentication step, use consistent domains where possible, and offer a trusted help route.

Security messages should support rather than shame users. "Your session expired. Sign in again through the library portal" is more useful than "Invalid error." Good error design helps distinguish a normal interruption from a suspicious request.

> **Ethical checkpoint.** A free chat widget permits evening reference service, but its vendor stores transcripts and device identifiers in another jurisdiction. What benefit, data flow, contract term, alternative, and user notice should be examined?

## Review questions

1. **Identification:** What is the difference between the internet and the World Wide Web?
2. **Identification:** Define bandwidth and latency. How can each affect a library website?
3. **Short answer:** What role does DNS play after a user enters a domain name?
4. **Short answer:** Identify the URL parts in https://archive.example.ph/items/42?view=full#description.
5. **Short answer:** Distinguish the roles of HTML, CSS, and JavaScript.
6. **Applied analysis:** A catalogue works on staff computers but not from homes. Give three plausible causes from different layers and one test for each.
7. **Applied analysis:** After a repository migration, search results lead to 404 responses. What should the team have prepared, and what should it do now?
8. **Scenario:** A homepage loads 12 MB of poster images before showing hours. Evaluate access, performance, metadata, and maintenance.
9. **Scenario:** A browser reports an invalid certificate. Why should staff not tell users to bypass the warning?
10. **Synthesis:** Choose one component from @tbl-web-governance and explain how its maintenance connects to an LIS responsibility.

## Mini case: The multicampus library website

Eastern Bay University has three campuses and a growing distance-learning program. Each library maintains separate CMS pages. Labels differ: one says "OPAC," another "Books," and the third "Library Search." Hours are poster images, two from last semester. Database links work on the main campus but fail for many remote students. The website loads six tracking and social-media domains. A screen-reader user reports repeated unlabeled links. Communications controls templates, IT controls authentication and DNS, vendors control database platforms, and no office believes it owns the complete journey.

**Guide questions**

1. Map the technical and organizational components from homepage to authenticated database article. Where should evidence be collected?
2. Prioritize five improvements using impact, urgency, effort, accessibility, privacy, and institutional control.
3. Propose governance naming content owners, technical owners, review intervals, escalation paths, and two success measures.

## Further reading and tools

### Standards and documentation

* **RFC 1034, _Domain Names---Concepts and Facilities_.** Foundational DNS concepts [@mockapetris1987dns].
* **RFC 9110, _HTTP Semantics_.** HTTP methods, status codes, fields, and representations [@fielding2022http].
* **WHATWG HTML Living Standard.** The normative reference for HTML semantics [@whatwg2026html].
* **WCAG 2.2.** Testable web accessibility requirements [@w3cwcag22].
* **W3C mobile accessibility guidance.** Application of accessibility guidance to mobile contexts [@w3c2015mobile].

### LIS readings

* @emde2009usability demonstrates task-based testing with academic-library users.
* @billingham2014accessibility discusses improving academic-library access for people with disabilities.
* @vargas2026homepage connects homepage structure and aesthetics with usability and accessibility.

### Open and widely accessible tools

* **Browser developer tools:** inspect requests, status codes, structure, responsive layouts, and accessibility.
* **W3C Nu HTML Checker:** identify HTML conformance errors and warnings.
* **WAVE:** support human accessibility review by marking potential issues.
* **Lighthouse:** generate performance and accessibility diagnostics.
* **WebPageTest:** examine loading under varied devices and connections.
* **diagrams.net:** draw request paths and responsibility maps.
* **Visual Studio Code or Positron:** edit HTML with syntax support.

### Chapter 2 bridge

Web pages and network requests produce and consume structured information. The next chapter moves below the interface to databases, where records, fields, identifiers, relationships, validation rules, and queries organize library data. Keep your URL analysis and request trace: both already contain structured components that can be modeled as data.
