---
title: "# Detailed Textbook Outline {.unnumbered}"
number-sections: false
---

```{r}
#| label: setup
#| include: false

# Output-aware tables: tabularx for PDF, Pandoc pipe tables for DOCX,
# and escaped HTML tables for HTML.
latex_escape <- function(x) {
  x <- gsub("\\", "<<<BACKSLASH>>>", as.character(x), fixed = TRUE)
  x <- gsub("([#$%&_{}])", "\\\\\\1", x, perl = TRUE)
  x <- gsub("~", "\\textasciitilde{}", x, fixed = TRUE)
  gsub("<<<BACKSLASH>>>", "\\textbackslash{}", x, fixed = TRUE)
}

ict_table <- function(data, caption, label = NULL, widths = NULL) {
  data <- as.data.frame(data, stringsAsFactors = FALSE)
  names(data) <- gsub("_", " ", names(data), fixed = TRUE)
  anchor <- if (!is.null(label)) paste0("{#tbl-", label, "}") else ""

  if (knitr::is_latex_output()) {
    align <- paste(rep("Y", ncol(data)), collapse = "")
    header <- paste(vapply(names(data), latex_escape, character(1)), collapse = " & ")
    rows <- apply(data, 1, function(row) {
      paste(vapply(row, latex_escape, character(1)), collapse = " & ")
    })
    if (grepl("^Key concepts", caption)) {
      column_weights <- if (is.null(widths) || length(widths) != ncol(data)) {
        rep(1, ncol(data))
      } else {
        widths
      }
      column_widths <- 0.94 * column_weights / sum(column_weights)
      long_align <- paste(
        sprintf(">{\\raggedright\\arraybackslash}p{%.3f\\textwidth}", column_widths),
        collapse = ""
      )
      out <- c(
        sprintf("\\begin{longtable}{%s}", long_align),
        sprintf("\\caption{%s}\\label{tbl-%s}\\\\", latex_escape(caption), label),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endfirsthead",
        sprintf("\\multicolumn{%d}{c}{\\tablename\\ \\thetable{} -- continued} \\\\" , ncol(data)),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endhead",
        sprintf("\\midrule \\multicolumn{%d}{r}{Continued on next page} \\\\" , ncol(data)),
        "\\endfoot", "\\bottomrule", "\\endlastfoot",
        paste0(rows, " \\\\"), "\\end{longtable}"
      )
    } else {
      out <- c(
        sprintf("\\begin{tabularx}{\\textwidth}{%s}", align),
        "\\toprule", paste0(header, " \\\\"), "\\midrule",
        paste0(rows, " \\\\"), "\\bottomrule", "\\end{tabularx}"
      )
    }
  } else if (knitr::is_html_output()) {
    if (grepl("^Key concepts", caption)) {
      out <- knitr::kable(data, format = "html", caption = caption, escape = TRUE,
                          table.attr = if (is.null(label)) "" else sprintf("id='tbl-%s'", label))
    } else {
      out <- knitr::kable(data, format = "html", escape = TRUE)
    }
  } else {
    if (grepl("^Key concepts", caption)) {
      out <- c(paste0(": ", caption, " ", anchor), "", knitr::kable(data, format = "pipe"))
    } else {
      out <- knitr::kable(data, format = "pipe")
    }
  }
  knitr::asis_output(paste(out, collapse = "\n"))
}
```

## Chapter 1. ICT in LIS as Service Infrastructure

**Summary.** The opening chapter reframes ICT from “computers and software” to a service infrastructure made of people, data, metadata, standards, workflows, policies, networks, and interfaces. Students follow a library request through a layered system, distinguish digitization from digital transformation, and examine how technical choices affect access, privacy, reliability, and professional values.

**Learning outcomes.** By the end of the chapter, students will be able to:

1. Define ICT, information infrastructure, system, platform, interoperability, and sociotechnical system in LIS terms.
2. Map the people, data, technology, policy, and workflow components of an information service.
3. Trace one user request through interface, application, metadata, data, and infrastructure layers.
4. Distinguish digitization, digitalization, automation, and digital transformation.
5. Identify likely service failures, affected stakeholders, and basic safeguards.
6. Evaluate an ICT decision using access, privacy, inclusion, sustainability, and accountability criteria.

**Major concepts.** ICT; infrastructure; sociotechnical system; information system; data; metadata; hardware; software; network; platform; workflow; standard; interoperability; automation; digitization; digitalization; digital transformation; availability; stakeholder; digital divide.

**Practical activity.** Conduct a service infrastructure audit of a familiar catalogue, repository, circulation service, or reference channel using a five-part worksheet.

**Expected output.** A one-page service map, a failure-and-impact table, and a 150-word reflection proposing one realistic improvement.

## Chapter 2. Networking, Web Basics, and Library Websites

**Summary.** Students investigate what happens between entering a URL and receiving a web page. They connect clients, servers, IP addresses, domain names, DNS, HTTP/HTTPS, browsers, HTML, CSS, hosting, and content management systems to library websites and remote access services. Website governance and low-bandwidth design are treated as service concerns.

**Learning outcomes.** Students will be able to:

1. Diagram a basic client-server transaction.
2. Explain the roles of IP addresses, DNS, URLs, HTTP, and HTTPS.
3. Inspect the structure of a simple HTML document.
4. Evaluate a library website for findability, trust, mobile use, and connection constraints.
5. Publish a small standards-based web page with semantic headings and links.

**Major concepts.** LAN; WAN; internet; web; protocol; packet; bandwidth; latency; client; server; DNS; URL; HTTP; HTTPS; HTML; CSS; browser; hosting; CMS; responsive design.

**Practical activity.** Use browser developer tools and a text editor to inspect and revise a small library-hours page.

**Expected output.** An annotated request-path diagram, valid HTML file, and short website critique. This lab contributes to the short-lab component and prepares students for the mobile and accessibility critique in Chapter 9.

## Chapter 3. Databases for Library Data

**Summary.** This chapter explains why structured storage is necessary for collections, users, transactions, and authorities. Students move from a messy spreadsheet to a simple relational model and learn tables, fields, records, keys, relationships, validation, normalization, queries, backups, and access control without assuming prior programming experience.

**Learning outcomes.** Students will be able to:

1. Distinguish flat files, spreadsheets, and relational databases.
2. Identify entities, attributes, primary keys, and relationships in an LIS scenario.
3. Create a small normalized database schema.
4. Enter validated data and run basic `SELECT`, `WHERE`, and `ORDER BY` queries.
5. Explain privacy and retention risks in patron and transaction data.

**Major concepts.** Database; DBMS; table; row; column; entity; attribute; schema; primary key; foreign key; relationship; normalization; SQL; query; constraint; transaction; backup.

**Practical activity.** Build a three-table SQLite database for a community reading-room collection, creators, and loans using DB Browser for SQLite.

**Expected output.** Entity-relationship diagram, `.sqlite` file, five queries with results, and a brief data-minimization note. This can begin the database or API mini-project.

## Chapter 4. Metadata and Cataloguing

**Summary.** Metadata is presented as structured, governed description that enables identification, selection, access, management, rights control, and preservation. Students compare descriptive, administrative, structural, technical, and preservation metadata; examine elements, values, schemas, encoding, content standards, identifiers, authority control, and application profiles; and connect MARC 21, RDA, Dublin Core, and emerging linked-data models.

**Learning outcomes.** Students will be able to:

1. Classify metadata by function.
2. Distinguish element sets, content standards, encoding formats, and value vocabularies.
3. Create a small Dublin Core record using controlled values.
4. Explain the purpose of identifiers and authority control.
5. Detect and correct common metadata quality problems.
6. Discuss how description can include, exclude, or misrepresent communities.

**Major concepts.** Metadata; element; value; schema; application profile; MARC 21; RDA; Dublin Core; identifier; authority control; controlled vocabulary; subject access; provenance; metadata quality; interoperability.

**Practical activity.** Describe a digitized local-history item in a shared metadata template, then exchange records for quality review.

**Expected output.** Completed metadata record, validation checklist, and revision log. This is the core metadata and cataloguing assessment.

## Chapter 5. Digital Libraries, Repositories, and Content Platforms

**Summary.** Students compare digital libraries, institutional repositories, digital archives, digital exhibits, and commercial content platforms by purpose, content model, workflow, governance, and preservation responsibility. Ingest, storage, metadata, access, rights, identifiers, and platform selection are tied to collection policy and institutional capacity.

**Learning outcomes.** Students will be able to:

1. Differentiate major kinds of digital content systems.
2. Diagram a basic ingest-to-access workflow.
3. Compare repository platforms using service and stewardship criteria.
4. Prepare files and metadata for a small digital collection.
5. Identify rights, consent, cultural sensitivity, and sustainability risks.

**Major concepts.** Digital library; repository; collection management system; CMS; ingest; submission information package; access copy; master file; OCR; persistent identifier; embargo; takedown; platform governance.

**Practical activity.** Create a three-item prototype collection in Omeka S or a structured local simulation.

**Expected output.** Collection statement, item records, rights labels, screenshots or export, and workflow diagram.

## Chapter 6. Information Retrieval and Discovery Systems

**Summary.** This chapter follows the path from a query to a ranked result set. It explains indexes, fields, tokenization, Boolean logic, phrase searching, facets, relevance ranking, precision, recall, federated search, link resolution, discovery layers, and evaluation. Students learn that retrieval quality depends as much on metadata and interface choices as on algorithms.

**Learning outcomes.** Students will be able to:

1. Explain how an index supports retrieval.
2. Construct fielded, Boolean, phrase, and faceted searches.
3. Calculate and interpret precision and recall in a small example.
4. Diagnose retrieval problems involving metadata, indexing, ranking, or interface design.
5. Compare catalogue, database, repository, and discovery-layer results.

**Major concepts.** Information retrieval; corpus; index; token; query; Boolean operator; fielded search; facet; relevance; ranking; precision; recall; stemming; synonym; federated search; discovery layer; link resolver.

**Practical activity.** Run the same information need through three discovery systems and analyze the first ten results.

**Expected output.** Search log, relevance judgments, simple precision comparison, and discovery-improvement recommendation.

## Chapter 7. Programming for LIS and Automation

**Summary.** Programming is introduced as a method for expressing repeatable decisions, not as an end in itself. Using R, students work with variables, vectors, data frames, conditions, loops or vectorized operations, functions, packages, errors, and reproducible scripts to clean filenames or metadata and generate a report.

**Learning outcomes.** Students will be able to:

1. Read and modify a short R script.
2. Represent metadata in vectors and data frames.
3. Apply conditions and functions to automate a repetitive task.
4. Import, clean, and export a small CSV file.
5. Test results and document assumptions.
6. Identify tasks that should remain under human review.

**Major concepts.** Algorithm; program; script; variable; data type; vector; data frame; condition; iteration; function; package; input/output; exception; debugging; reproducibility; automation.

**Practical activity.** Use R and `tidyverse` packages to standardize dates, trim spaces, flag missing titles, and export a metadata quality report.

**Expected output.** Commented `.R` or `.qmd` script, cleaned CSV, exception report, and verification checklist.

## Chapter 8. APIs, Linked Data, and the Semantic Web

**Summary.** Students learn how systems exchange structured data through APIs and how linked data expresses relationships using web identifiers and triples. Requests, responses, endpoints, parameters, status codes, JSON, XML, RDF, URIs, vocabularies, and SPARQL are introduced through library authority and bibliographic data.

**Learning outcomes.** Students will be able to:

1. Describe a basic API request-response exchange.
2. Read nested JSON and identify relevant fields.
3. Retrieve and transform a small public API response in R.
4. Express a bibliographic relationship as RDF-like subject-predicate-object triples.
5. Explain the benefits and risks of identifiers and linked open data.

**Major concepts.** API; endpoint; request; response; parameter; authentication; status code; JSON; XML; URI; RDF; triple; vocabulary; ontology; knowledge graph; SPARQL; linked open data.

**Practical activity.** Query an open books or authority API, select five fields, and build a small subject-guide table with source attribution.

**Expected output.** Reproducible API script, saved raw response, cleaned dataset, data dictionary, and attribution note. This completes the database or API mini-project.

## Chapter 9. UX, Accessibility, and Mobile Library Services

**Summary.** User experience (UX) is treated as evidence-based service design. Students examine user research, task analysis, information architecture, usability, accessibility, responsive and mobile-first design, assistive technologies, plain language, and WCAG. They consider constrained devices, prepaid data, intermittent connectivity, and multilingual interaction in Southeast Asian contexts.

**Learning outcomes.** Students will be able to:

1. Distinguish usability, UX, and accessibility.
2. Create a task-based usability test.
3. Evaluate a service against selected WCAG 2.2 criteria.
4. Identify barriers in navigation, content, forms, and mobile layouts.
5. Prioritize changes by user impact and implementation effort.

**Major concepts.** UX; usability; accessibility; inclusive design; user research; persona; task; information architecture; responsive design; mobile first; assistive technology; alt text; keyboard access; contrast; WCAG.

**Practical activity.** Test a library discovery task on desktop and mobile with an accessibility checklist and one peer participant.

**Expected output.** Test protocol, observation notes, annotated screens, severity-ranked findings, and redesign sketch. This is the UX or discovery critique.

## Chapter 10. Digital Preservation and Cloud Services

**Summary.** Preservation is distinguished from storage, backup, and digitization. Students examine threats, file formats, checksums, fixity, redundancy, migration, emulation, preservation metadata, storage tiers, cloud service models, vendor dependence, cost, jurisdiction, and exit planning through the lifecycle of a digital object.

**Learning outcomes.** Students will be able to:

1. Distinguish backup, storage, digitization, and digital preservation.
2. Identify common threats to digital continuity.
3. Generate and verify checksums.
4. Propose a basic 3-2-1 storage and monitoring plan.
5. Compare local and cloud options using cost, control, security, jurisdiction, and exit criteria.

**Major concepts.** Digital preservation; bitstream; fixity; checksum; integrity; authenticity; redundancy; 3-2-1 rule; migration; emulation; obsolescence; preservation metadata; IaaS; PaaS; SaaS; service-level agreement; vendor lock-in.

**Practical activity.** Create a small preservation package with inventory, checksums, format-risk notes, three copies, and a recovery test.

**Expected output.** Package manifest, fixity log, storage diagram, and preservation action plan.

## Chapter 11. Cybersecurity, Law, Ethics, and Project Planning

**Summary.** The chapter integrates risk management with professional responsibility. Students study confidentiality, integrity, availability, threats, vulnerabilities, controls, authentication, authorization, least privilege, phishing, malware, incident response, copyright, privacy, intellectual freedom, procurement, scope, stakeholders, schedules, budgets, and project risks.

**Learning outcomes.** Students will be able to:

1. Conduct a basic asset-threat-vulnerability-control analysis.
2. Recommend proportionate administrative, technical, and physical controls.
3. Apply privacy, copyright, and intellectual-freedom questions to an ICT project.
4. Recognize common social-engineering attempts and report them appropriately.
5. Draft a project charter with scope, deliverables, roles, timeline, and risk register.
6. Explain when legal, IT, records, or community expertise is required.

**Major concepts.** Confidentiality; integrity; availability; asset; threat; vulnerability; risk; control; authentication; authorization; least privilege; encryption; incident; privacy impact assessment; copyright; fair use; intellectual freedom; project charter; scope; milestone; risk register.

**Practical activity.** Prepare a security, ethics, and project-start brief for a proposed small repository.

**Expected output.** Threat model, privacy checklist, rights workflow, one-page project charter, and risk register. This forms the ethics and preservation brief with Chapter 10.

## Chapter 12. Data Literacy and Capstone Integration

**Summary.** Students bring the course together by framing questions, evaluating data provenance and quality, cleaning and summarizing data, choosing honest visualizations, documenting limitations, and turning evidence into service decisions. Capstone planning emphasizes users, metadata, interoperability, ethics, sustainability, evaluation, and communication.

**Learning outcomes.** Students will be able to:

1. Formulate an answerable service question.
2. Assess a dataset for provenance, completeness, validity, consistency, and bias.
3. Produce a reproducible descriptive analysis and accessible visualization in R.
4. Interpret results without overstating what the data can support.
5. Integrate technical, metadata, user, ethical, and sustainability requirements in a capstone.
6. Present and defend a prototype or policy deliverable to mixed stakeholders.

**Major concepts.** Data literacy; provenance; data quality; tidy data; missingness; descriptive statistics; denominator; visualization; dashboard; inference; reproducibility; documentation; evaluation metric; theory of change; capstone.

**Practical activity.** Analyze a small anonymized service dataset in R and use the evidence to refine a capstone proposal.

**Expected output.** `.qmd` analysis, cleaned data, one table, one accessible chart, limitations statement, capstone artifact, presentation, and individual reflection.

### Capstone options

Students choose one of the six options in @tbl-capstones. Each option demands an integrated response rather than a single technical artifact: the assessed evidence includes service rationale, metadata or data structures, user needs, ethical analysis, implementation documentation, evaluation, and sustainability.

```{r}
#| label: tbl-capstones
#| tbl-cap: "Capstone options and required integration"
#| results: asis
capstones <- data.frame(
  Option = c("Campus digital exhibit", "API-powered subject guide",
             "Discovery improvement study", "Small repository policy pack",
             "Metadata cleanup and enrichment project", "Mobile service redesign"),
  Core_artifact = c("Curated exhibit prototype", "Working guide and API workflow",
                    "Evidence-based findings and redesign", "Policy and workflow set",
                    "Cleaned records and enrichment log", "Tested mobile prototype"),
  Required_integration = c("Selection, description, rights, access, preservation",
                           "API, data transformation, attribution, accessibility",
                           "Retrieval, metadata, UX testing, evaluation",
                           "Governance, metadata, preservation, security",
                           "Quality rules, automation, authority work, audit trail",
                           "User research, accessibility, privacy, low-bandwidth design")
)
ict_table(capstones, "Capstone options and required integration", "capstones", c(1.1, 1.4, 2))
```

### Capstone rubric

@tbl-capstone-rubric makes the capstone expectations visible early. Technical execution is important, but it receives only one portion of the grade; an impressive prototype that is inaccessible, poorly described, undocumented, or unsustainable does not meet LIS professional standards.

```{r}
#| label: tbl-capstone-rubric
#| tbl-cap: "Common capstone rubric"
#| results: asis
capstone_rubric <- data.frame(
  Criterion = c("Service problem and users", "Metadata and data quality", "Technical implementation",
                "Ethics, rights, privacy, and accessibility", "Evaluation and evidence",
                "Documentation, sustainability, and communication"),
  Weight = c("15%", "20%", "20%", "15%", "15%", "15%"),
  Strong_performance = c("Need and stakeholders are specific and evidence based",
                         "Structures and values are consistent, justified, and interoperable",
                         "Artifact works reliably; tests and limitations are documented",
                         "Risks are identified and addressed through concrete choices",
                         "Criteria and evidence support a defensible judgment",
                         "Another team could understand, maintain, and explain the work")
)
ict_table(capstone_rubric, "Common capstone rubric", "capstone-rubric", c(1.2, .5, 2.5))
```

### Recurring chapter design

Every full chapter follows the same learning rhythm: encounter a service problem, develop a vocabulary, construct a working model, practice with a small tool or dataset, examine an ethical tension, and transfer the learning to a case. This regularity reduces unnecessary cognitive load while allowing the technical content to grow more demanding.

Each lab also produces a portfolio object. By Week 12, the learner has maps, schemas, records, scripts, critiques, policies, and reflections from which to build a capstone. Short reflective prompts ask not only “Did it work?” but also “For whom did it work, under what assumptions, and who carries the risk when it fails?”
