---
title: "# Preface {.unnumbered}"
number-sections: false
---

```{r}
#| label: setup
#| include: false

# Output-aware tables: tabularx for PDF, Pandoc pipe tables for DOCX,
# and escaped HTML tables for HTML.
latex_escape <- function(x) {
  x <- gsub("\\", "<<<BACKSLASH>>>", as.character(x), fixed = TRUE)
  x <- gsub("([#$%&_{}])", "\\\\\\1", x, perl = TRUE)
  x <- gsub("~", "\\textasciitilde{}", x, fixed = TRUE)
  gsub("<<<BACKSLASH>>>", "\\textbackslash{}", x, fixed = TRUE)
}

ict_table <- function(data, caption, label = NULL, widths = NULL) {
  data <- as.data.frame(data, stringsAsFactors = FALSE)
  names(data) <- gsub("_", " ", names(data), fixed = TRUE)
  anchor <- if (!is.null(label)) paste0("{#tbl-", label, "}") else ""
  if (knitr::is_latex_output()) {
    align <- paste(rep("Y", ncol(data)), collapse = "")
    header <- paste(vapply(names(data), latex_escape, character(1)), collapse = " & ")
    rows <- apply(data, 1, function(row) {
      paste(vapply(row, latex_escape, character(1)), collapse = " & ")
    })
    if (grepl("^Key concepts", caption)) {
      column_weights <- if (is.null(widths) || length(widths) != ncol(data)) {
        rep(1, ncol(data))
      } else {
        widths
      }
      column_widths <- 0.94 * column_weights / sum(column_weights)
      long_align <- paste(
        sprintf(">{\\raggedright\\arraybackslash}p{%.3f\\textwidth}", column_widths),
        collapse = ""
      )
      out <- c(
        sprintf("\\begin{longtable}{%s}", long_align),
        sprintf("\\caption{%s}\\label{tbl-%s}\\\\", latex_escape(caption), label),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endfirsthead",
        sprintf("\\multicolumn{%d}{c}{\\tablename\\ \\thetable{} -- continued} \\\\" , ncol(data)),
        "\\toprule", paste0(header, " \\\\"), "\\midrule", "\\endhead",
        sprintf("\\midrule \\multicolumn{%d}{r}{Continued on next page} \\\\" , ncol(data)),
        "\\endfoot", "\\bottomrule", "\\endlastfoot",
        paste0(rows, " \\\\"), "\\end{longtable}"
      )
    } else {
      out <- c(
        sprintf("\\begin{tabularx}{\\textwidth}{%s}", align),
        "\\toprule", paste0(header, " \\\\"), "\\midrule",
        paste0(rows, " \\\\"), "\\bottomrule", "\\end{tabularx}"
      )
    }
  } else if (knitr::is_html_output()) {
    if (grepl("^Key concepts", caption)) {
      out <- knitr::kable(data, format = "html", caption = caption, escape = TRUE,
                          table.attr = if (is.null(label)) "" else sprintf("id='tbl-%s'", label))
    } else {
      out <- knitr::kable(data, format = "html", escape = TRUE)
    }
  } else {
    if (grepl("^Key concepts", caption)) {
      out <- c(paste0(": ", caption, " ", anchor), "", knitr::kable(data, format = "pipe"))
    } else {
      out <- knitr::kable(data, format = "pipe")
    }
  }
  knitr::asis_output(paste(out, collapse = "\n"))
}
```

Information and communication technology (ICT) in Library and Information Science (LIS) is not merely a list of office applications. It is the sociotechnical infrastructure through which information organizations acquire, describe, connect, retrieve, deliver, preserve, protect, govern, and evaluate information. A catalogue search, a digitized newspaper, an archival finding aid, a circulation notice, and a research-data dashboard all depend on interactions among people, policies, metadata, standards, devices, networks, and software.

This book is intended for undergraduate LIS students with little or no formal technical background. It begins with service problems and moves toward technical ideas only as those ideas become useful. Metadata and knowledge organization form the book's spine. Hands-on work starts in Week 1, ethical questions recur throughout, and each technical topic is tied to professional responsibilities in libraries, archives, records centers, museums, and related information organizations.

The present volume contains a detailed plan and complete text for all twelve chapters. Citations use Pandoc/Quarto keys such as [@star1996steps] and narrative forms such as @borgman2007scholarship. The bibliography is maintained separately in `references.bib` so that sources can be reused and checked as the manuscript is revised.

## Course architecture and assessment

The six-part architecture moves from foundations to description, services, code, stewardship, and synthesis. @tbl-course-map shows how the twelve chapters create that progression and where the course assessments fit. The table matters because assessment evidence is distributed: students repeatedly practice small skills before integrating them in a capstone.

```{r}
#| label: tbl-course-map
#| tbl-cap: "Course architecture and assessment alignment"
#| results: asis
course_map <- data.frame(
  Architecture = c("Foundations of ICT in LIS", "Data Structures and Description",
                   "Discovery and Digital Services", "Coding and Web Data",
                   "Stewardship, Risk, and Governance", "Data Practice and Synthesis"),
  Chapters = c("1-2", "3-4", "5-6 and 9", "7-8", "10-11", "12"),
  Principal_evidence = c("Short labs and quizzes", "Metadata and cataloguing work",
                         "UX or discovery critique", "Database or API mini-project",
                         "Ethics and preservation brief", "Capstone"),
  Weight = c("20%", "15%", "10%", "20%", "10%", "25%")
)
ict_table(course_map, "Course architecture and assessment alignment", "course-map", c(1.5, .7, 1.4, .5))
```

The sequence is cumulative rather than strictly linear. For example, privacy first appears as a design consideration in Chapter 1, becomes a data issue in Chapter 3, and returns as a legal and governance obligation in Chapter 11. Likewise, metadata introduced in Chapter 1 is formalized in Chapter 4 and then reused in repositories, discovery, linked data, and preservation.
